\dnest~Manual

### 2016-04-06 doc/manual.tex
PYTHON_PATH
Some editing

If you find \dnest~useful in a research project,
please feel free to cite the paper.
Throughout this manual I will assume that you have read the paper and have a
reasonable understanding of how the algorithm works, as well as some
understanding of general Nested Sampling concepts.\\

\section{Installation}
Please note that I have only ever compiled \dnest~on GNU/Linux,
using the GNU C++ compiler ({\tt g++}). However, I expect it to be
straightforward to compile it on any other Unix-like operating system such as Mac
OS X or FreeBSD. It should be possible, albeit probably more tricky,
to compile \dnest~in Microsoft Windows. If you want to try,
the MinGW compiler is
probably your best bet. I'd be interested to hear from anyone who
has tried this. I would also be interested to hear if anyone has
compiled \dnest~with a different compiler, such as the Intel C++
compiler.\\

\subsection{Dependencies}
One of the issues with DNest3 which inspired a new version was that many users
found it annoying to compile on different operating systems because of the
dependencies. For that reason, \dnest~now has much less dependencies.
The C++ part of the code is written in pure C++11 and does not require any
additional libraries --- you just have to have a modern C++ compiler that
supports the C++11 standard.

The Python postprocessing scripts use the Python packages
{\tt numpy} and {\tt matplotlib}. They work with both Python 2 or 3.

\subsection{Compiling}
Go into the {\tt code} directory and run {\tt make}. Everything should compile.
To
verify that the compilation was successful,
go into the code/Examples/SpikeSlab directory
and see if an executable file called {\tt main} has been created. If you
execute this file, you should see a lot of output about ``levels'' and
``particles'' (see Figure~\ref{fig:output} for example output).
You will need to manually terminate this process by using
Control+C, or whatever works in your operating system. If you saw this output,
\dnest~was compiled successfully.\\

\section{Running DNest}
As part of the compilation process, several examples will have been compiled.
There is one executable file for each example: \dnest~does not have a global
executable. The
following discussion will involve running the ``SpikeSlab'' example, which is
the example used in the paper.\\

To run the SpikeSlab example, go into the code/Examples/Spikeslab
directory and execute {\tt main}. You should see some output that looks like
that shown in Figure~\ref{fig:output}.
Note that you must terminate the process manually by pressing Ctrl+C
or in another way. If you do not terminate the process yourself, it will run
forever. There is no definite rule for how long you should run \dnest.
It depends on the difficulty of your problem, the number of posterior samples you
want, and the accuracy you want for the marginal likelihood.\\

@@

The executable {\tt main} is responsible for the exploration part of the
DNS algorithm (i.e. running the MCMC chain, building
levels, and then exploring all the levels). It creates three output text files,
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.\\

The first output
file, {\tt sample.txt}, contains a sampling of parameter values that
represents the {\it mixture of constrained priors} (the target distribution
used in DNS), {\bf not} the
posterior distribution. Each line of {\tt sample.txt} represents a point in
parameter space. In the SpikeSlab example, there are 20 parameters, so there
will be 20 columns in {\tt sample.txt}.
Each time a point is saved to {\tt sample.txt}, \dnest~prints
the message ``Saving a particle to disk. N = ...''.\\

@@

The third output file, {\tt levels.txt}, contains information about the levels
that were built during the run. The first column has estimates of the $\log(X)$
values of the levels, i.e. how compressed they are relative to the prior, in
units of nats. The second column contains the log likelihoods of the levels.
The first level, with a $\log(X)$ value of 0 and a log likelihood of
$-10^{300}$ (basically ``minus infinity''), is simply the prior. The third
column has the ``tiebreaker'' values for the levels, which again are not
particularly useful unless your problem has likelihood plateaus. The fourth
and fifth columns are the number of accepted proposals and the total number
of proposals that have occurred within each level, which are useful for
monitoring the Metropolis acceptance ratio as a function of level.
The final two columns, called ``exceeds'', and ``visits'', are used to refine
the estimates of the level compressions (and hence the $\log(X)$ values of
the levels in column 1), as discussed in Section 3 of the
paper. The visits column counts the number of times a level (level $j$, say)
has been visited, but only starts counting after the next level ($j+1$) has been created. The exceeds column counts the number of times a particle that was
in level $j$ had a likelihood that exceeded that of level $j+1$.
\\

>>>>

If you find \dnest~useful in a research project,
please cite the paper.
Throughout this manual I will assume that you have read the paper and have a
reasonable understanding of how the algorithm works, as well as some
understanding of general Nested Sampling concepts.\\

\section{Installation}
I have only ever compiled \dnest~on GNU/Linux,
using the GNU C++ compiler ({\tt g++}). However, I expect it to be
straightforward to compile it on any other Unix-like operating system such as Mac
OS X or FreeBSD.\\

It should be possible, but probably more tricky,
to compile \dnest~in Microsoft Windows. If you want to try,
the MinGW compiler is
probably your best bet. I'd be interested to hear from anyone who
has tried this, and any problems that arose.
I would also be interested to hear if anyone has
compiled \dnest~with a different compiler, such as the Intel C++
compiler.\\

\subsection{Dependencies}
One of the issues with DNest3 (an older implementation) which inspired a rewrite
was that many users
found it annoying to compile on different operating systems because of the
dependencies. For that reason, \dnest~now has much less dependencies.
The C++ part of the code is written in pure C++11 and does not require any
additional libraries --- you just have to have a modern C++ compiler that
supports the C++11 standard.

The Python postprocessing scripts use the Python packages
{\tt NumPy}, {\tt matplotlib}, {\tt Pandas}, and {\tt Cython}.
They should work with both Python 2 or 3.

\subsection{Compiling}
Go into the {\tt code} directory and run {\tt make}. Everything should compile.
To
verify that the compilation was successful,
go into the {\tt code/Examples/SpikeSlab} directory
and see if an executable file called {\tt main} has been created. If you
execute this file, you should see a lot of output about ``levels'' and
``particles'' (see Figure~\ref{fig:output} for example output).
You will need to manually terminate this process by using
Control+C, or whatever works in your operating system. If you saw this output,
the C++ parts of \dnest~were compiled successfully.\\

Next, you'll need to install the {\tt dnest4} Python package so you can use
the postprocessing scripts, and other features. Go into the
{\tt python} directory and run {\tt python setup.py install} or similar.

\section{Running DNest4}
As part of the compilation process, several examples will have been compiled.
There is one executable file for each example: \dnest~does not have a global
executable. The
following discussion will involve running the ``SpikeSlab'' example, which is
the example used in the paper.\\

To run the SpikeSlab example, go into the {\tt code/Examples/Spikeslab}
directory and execute {\tt main}. You should see some output that looks like
that shown in Figure~\ref{fig:output}.
You must terminate the process manually by pressing Ctrl+C
or in another way. If you do not terminate the process yourself, it will run
forever. There is no definite rule for how long you should run \dnest. Like
many MCMC methods, it depends on how hard your problem is, the number of
posterior samples you
want, and the accuracy you want for the marginal likelihood or other
quantities you care about.\\

@@

The executable {\tt main} is responsible for the exploration part of the
DNS algorithm (i.e. running the MCMC chain, building
levels, and then exploring all the levels). It creates three text output files,
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.\\

The first output
file, {\tt sample.txt}, contains a sampling of parameter values that
represents the {\it mixture of constrained priors} (the target distribution
used in DNS), {\bf not} the
posterior distribution. Each line of {\tt sample.txt} represents a point in
parameter space. In the SpikeSlab example, there are 20 parameters, so there
are 20 columns in {\tt sample.txt}.
Each time a point is saved to {\tt sample.txt}, \dnest~prints
the message ``Saving a particle to disk. N = ...''.\\

@@

The third output file, {\tt levels.txt}, contains information about the levels
that were built during the run. The first column has estimates of the $\log(X)$
values of the levels, i.e. how compressed they are relative to the prior, in
units of nats. For example, if a level has $\log(X) = -1.02$, its likelihood
value encloses $\exp(-1.02) \approx 36.1\%$ of the prior mass.\\

The second column contains the log likelihoods of the levels.
The first level, with a $\log(X)$ value of 0 and a log likelihood of
$-10^{308}$ (basically ``minus infinity''), is simply the prior. The third
column has the ``tiebreaker'' values for the levels, which again are not
particularly useful unless your problem has likelihood plateaus. The fourth
and fifth columns are the number of accepted proposals and the total number
of proposals that have occurred within each level, which are useful for
monitoring the Metropolis acceptance ratio as a function of level.
The final two columns, called ``exceeds'', and ``visits'', are used to refine
the estimates of the level compressions (and hence the $\log(X)$ values of
the levels in column 1), as discussed in Section 3 of the
paper. The visits column counts the number of times a level (level $j$, say)
has been visited, but only starts counting after the next level ($j+1$) has been created. The exceeds column counts the number of times a particle that was
in level $j$ had a likelihood that exceeded that of level $j+1$.
\\
